---
title: "labsession2"
output:
  pdf_document: default
  html_document: default
date: "2026-07-07"
editor_options:
  chunk_output_type: console
---

## R Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# setting the wd ----------------------------------------------------------
# setting it to the course folder

setwd(file.path(here::here(),"2026","DemogInR"))

# Packages
library(tidyverse)

# import functions from Tim for LT
source("https://github.com/timriffe/BSSD2026Module2/raw/refs/heads/master/02_lifetables.R")
```

# Today: lifetable and functions

Let's see some useful functions `cumsum()`, `cumprod`, `rev()`.

```{r}

x <- 1:10

cumsum(x)
cumprod(x)
rev(x)
rev(x) |> cumsum()
```

## reading the data

```{r}
mort <- read_csv(
  "https://github.com/timriffe/BSSD2026Module2/raw/refs/heads/master/data/ES_mort.csv.gz",
  show_col_types = FALSE
)
head(mort)
```


## step 1 - qx from mx

To calculate condition probabilities $q_x$, we start with $m_x$ and instrumental life table column denoted $a_x$.

To denote $a_x$, we use the following formula:

$$ q_x = \frac{m_x}{1 + (1-a_x) \cdot m_x}$$
For that, we need the $a_x$, and it has a meaning: it's the average years lived in the age interval by those that die in the interval.

-  Single-age groups: $a_x$ is around $\frac{1}{2}$

-  For infant mortality: we follow Adreev-Kingkade

```{r}
mx_to_ax <- function(mx, age, sex = "f"){
  ax <- case_when(
    age == 0 & sex == "m" & mx < 0.023 ~ 0.14929 - 1.99545 * mx,
    age == 0 & sex == "m" & mx >= 0.023 & mx < 0.08307 ~ 0.02832 + 3.26021 * mx,
    age == 0 & sex == "m" & mx >= 0.08307 ~ 0.29915,
    age == 0 & sex == "f" & mx < 0.01724 ~ 0.14903 - 2.05527 * mx,
    age == 0 & sex == "f" & mx >= 0.01724 & mx < 0.06891 ~ 0.04667 + 3.88089 * mx,
    age == 0 & sex == "f" & mx >= 0.06891 ~ 0.31411,
    age == max(age) ~ 1 / mx,
    TRUE ~ 0.5
  )
  return(ax)
}
```

Now we transform mx into qx using ax function

```{r}
# toy mx

mx <- mort |> 
  filter(sex == "female", year == 2000) |> 
  pull(mx)

mx_to_ax(mx, age = seq(min(mort$age),max(mort$age),1), sex = rep("f",111))
```

Creating function

```{r}
mx_to_qx <- function(mx, ax){
  qx <- mx / (1 + (1 - ax) * mx)
  qx <- if_else(qx > 1, 1, qx)
  return(qx)
}

# let's try it out
age = seq(min(mort$age),max(mort$age),1)
ax <- mx_to_ax(mx, age = age, sex = rep("f",111))

qx <- mx_to_qx(mx, ax)

```

# Step 2 - computing lx

First we need to compute px
```{r}
qx_to_px <- function(qx){
  px <- 1 - qx
  return(px)
}
```

Then we can compute lx as a product from zero to each age
```{r}
px_to_lx <- function(radix, age,px){
  lx <- case_when(
    age == 0 ~ radix,
    TRUE ~ radix * cumprod(px)
  )
  return(lx)
}
px <- qx_to_px(qx)

lx <- px_to_lx(radix = 1e5, age = age, px)
```

# step 3 - deriving ndx

$$d_x = \ell_x - \ell_{x+1} = \ell_x \cdot q_x$$
```{r}
lxqx_to_dx <- function(lx,qx){
  dx <- lx * qx
  return(dx)
}

dx <- lxqx_to_dx(lx,qx)
```

# step 4 - deriving Lx

```{r}
lxdx_to_Lx <- function(lx,dx,ax){
  Lx <- lx - (1 - ax) * dx
  return(Lx)
}

Lx <- lxdx_to_Lx(lx,dx,ax)
```

Lx can be thought of as lifetable exposure. It is a synthetic exposure underlying to the mortality rates we have.

# step 5 - deriving ex

We make an instrumental column Tx which is the total years lived from a specific age x.

```{r}
Lx_to_Tx <- function(Lx){
  Tx <- Lx |> 
    rev() |> 
    cumsum() |> 
    rev()
  return(Tx)
}
Tx <- Lx_to_Tx(Lx)
```

And to get the life expectancy at each age:

```{r}
Txlx_to_ex <- function(Tx, lx){
  ex <- Tx / lx
  return(ex)
}
ex <- Txlx_to_ex(Tx,lx)
```

# comparing estimates

```{r}
sum(lx) / 1e5
sum(dx * (age + ax)) / sum(dx)
```

# challenge

We have all of these functions. Now, we can create life table for female and male.

```{r}
lt <- mort |> 
  filter(sex != "total") |> 
  # replace and sme unexpected values for mortality
  arrange(sex, year, age) |> 
  tidyr::fill(
    mx,
    .direction = "down",
    .by = c(sex, year)
  ) |> 
  mutate(
    mx = if_else(mx == 0, .4, mx),
    ax = mx_to_ax(age = age, mx = mx, sex = first(sex)),
    qx = mx_to_qx(mx = mx, ax = ax),
    px = qx_to_px(qx = qx),
    lx = px_to_lx(radix = 1, px = px, age = age),
    dx = lx * qx,
    Lx = lxdx_to_Lx(lx = lx, dx = dx, ax = ax),
    Tx = Lx_to_Tx(Lx),
    ex = Txlx_to_ex(Tx, lx),
    .by = c(sex, year)
  )
```

## Doing in a Tim Riffe way

```{r}
ESlt <-
mort |> 
  filter(sex != "total") |> 
  arrange(sex, year, age) |> 
  # replaces NAs w value from lower age, only affects > 100
  tidyr::fill(mx,
              .direction = "down",
              .by = c(sex, year)) |> 
  mutate(
    # hack only affects ages > 100
      mx = if_else(mx == 0, .4, mx)) |> 
  group_by(sex, year) |> 
  group_modify(~lt_full(.x,.y,radix = 1)) |> 
  ungroup()
```

Graphs
```{r}
ESlt |> 
  filter(age == 0) |> 
  ggplot(aes(x = ex,y=qx,color = sex, alpha = year))+
  geom_point()
```

```{r}
ESlt |> 
  filter(age == 0) |> 
  ggplot(aes(x = year, y = ex, color = sex)) +
  geom_line()
```

