---
title: "labsession5 - mortality projection"
output:
  pdf_document: default
  html_document: default
date: "2026-07-10"
editor_options:
  chunk_output_type: console
---

## R Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# setting the wd ----------------------------------------------------------
# setting it to the course folder

setwd(file.path(here::here(),"2026","DemogInR"))

# Packages
library(tidyverse)

source("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/02_lifetables.R")

```

# Today: Predicting mortality

Read mortality data

```{r}
mx <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/data/ESmx.csv.gz")
mx |> pull(year) |> range() # female mortality ranging from 1950 to 2014
```

It easier to project female mortality due to less external mortality. Also, older adults mortality is easier.

# looking at the surface:

```{r}
mx |> 
  ggplot() +
  aes(x = year, y = age, z = log(mx)) +
  geom_contour_filled()
```

Let's look at it on the age dimension. And we can plot alpha on top of them

```{r}
alpha <- 
  mx |> 
  summarise(alphax = mean(log(mx)),.by = age)

mx |> 
  ggplot() +
  aes(x = age, y = mx) +
  geom_line(aes(color = year, group = year)) +
  scale_y_log10() +
  geom_line(data = alpha,
            aes(x = age, y= exp(alphax)),
            color = "red3",
            linetype = "dashed",
            linewidth = 1.2)
```

Alpha represents a centroid of the log mortality rates.

# get the matrix of centered log mortality

```{r}
clmx <- 
  mx |> 
  left_join(alpha, by = join_by(age)) |> 
  mutate(lmx = log(mx),
         clmx = lmx - alphax)
```
Examine the pattern

```{r}
clmx |> 
  ggplot() +
  aes(x = age, y = clmx, z = year, color = year, group = year) +
  geom_line()
```

Turn CLMX into a matrix following the Lexis format. We spread years over the columns, then make the age column into row names and finally coerce to be a matrix:

```{r}
M <- clmx |> 
  select(year, age, clmx) |> 
  pivot_wider(names_from = year,
              values_from = clmx) |> 
  column_to_rownames("age") |> 
  as.matrix()
```

Now we can SVD this matrix:

```{r}
svdM <- svd(M)
d <- svdM$d
U <- svdM$u
V <- svdM$v

# putting all the pieces together to recriate the original matrix

(U %*% diag(d) %*% t(V)) - M
```

This is a list of three elements. And we have to pick the most important information.

```{r}
betax <- U[,1] / sum(U[,1])

kappat <- V[,1] * d[1] * sum(U[,1])

plot(kappat)
```

Get drift parameters and do the forecasting and predict the surface we started with.

```{r}
years <- mx |> pull(year) |> unique() |> sort()

drift <-(kappat[length(kappat)] - kappat[1]) / ((2014 - 1950) - 1)

horizon <- -64:30
kappat_hat <- horizon * drift + kappat[length(kappat)]

```

Build tibbles and put the pieces together

```{r}
kappa <- 
  tibble(
    year = 2015 + horizon,
    kappat = kappat_hat
  )

betat <- tibble(age = 0:100, betax)

lc <- cross_join(kappa,betat) |> 
  left_join(alpha, by = join_by(age)) |> 
  mutate(mx = exp(alphax + betax * kappat))

```

Plotting the surface

```{r}
lc |> 
  ggplot() +
  aes(x = year, y = age, z = log(mx)) +
  geom_contour_filled()
```

Potting the rates by age
```{r}
lc |> 
  filter(year > 2014) |> 
  ggplot() +
  aes(x = age, y =mx, color = year, group = year) +
  geom_line() +
  scale_y_log10()
```

# Calculate the lifetables

```{r}
LTlc <- lc |> 
  mutate(sex = "female") |>
  group_by(year, sex) |>
  group_modify(~lt_full(.x,.y,radix = 1)) |> 
  ungroup()
```

Plotting e0 over time
```{r}
LTlc |> 
  filter(age == 0) |> 
  ggplot() +
  aes(x = year, y = ex) +
  geom_line()

```
Contrasting with the empirical LE

```{r}
mx |> 
  mutate(sex = "female") |>
  group_by(year, sex) |>
  group_modify(~lt_full(.x,.y,radix = 1)) |> 
  ungroup() |> 
  filter(age == 0) |> 
  ggplot() +
  aes(x = year, y = ex) +
  geom_point() +
  geom_line(
    data = LTlc |> filter(age == 0)
  )
```

