---
title: "labsession - workshop on Life Course Analysis (Christian Dudel)"
author: "tca"
format: html
editor_options: 
  chunk_output_type: console
---

# Package

We can find all info on github for the whole documentation, [click here](https://github.com/christiandudel/dtms).

```{r}
library(pacman)
p_load(dtms, tidyverse)
```

# Data and some miscelaneous stuffs in the package

We will start using simulated data into the package. This is a panel data in the long format.
```{r}
head(simpledata)
help(simpledata)

# exploring data structure
simpledata$time |> unique()
simpledata$id |> unique() |> length()
simpledata$state |> unique()
```

# Define model

We will setup the multistate model using the package `dtms`.

First we define the model using `dtms()`

```{r}
simple <- dtms(transient = c("A","B"),
               absorbing = "X",
               timescale = 0:19)
```

Transforming model into a transition format.

```{r}
estdata <- dtms_format(data = simpledata,
                       dtms = simple,
                       idvar = "id",
                       timevar = "time",
                       statevar = "state")
```

Having a summary of the data in the transition format
```{r}
estdata |> summary()
```

There are some stuffs that we have to get rid of transitions we do not want to have there. For that, there is a function for it.

```{r}
estdata <- dtms_clean(data = estdata,
                      dtms = simple)
estdata |> summary()
```

To deal with censoring in the data, we can also use a function for it.

```{r}
dtms_censoring(data = estdata,
               dtms = simple)
```

# Parametric models for transition probabilities

Let's fit a model using previous transition data

```{r}
fit <- dtms_fit(data = estdata,
                controls = "time")
fit |> summary()
```

To predict probabilities

```{r}
probs <- dtms_transitions(model = fit,
                          dtms = simple,
                          controls = list(time=0:19))

probs |> head()
probs |> summary()
estdata |> summary()
```

Let's check how it looks like now.

```{r}
# rbase way

# probs |> plot(dtms=simple)

# in a ggplot way

probs |> 
  dtms_simplify() |> 
  ggplot() +
  aes(x=time,y=P,color=to) +
  geom_line() +
  facet_wrap(~ from)
```

# Estimation

We can easily have the state expectancy using

```{r}

# state expectancy at starting state
dtms_expectancy(probs = probs,
                dtms = simple)

# unconditional state expectancy

s <- dtms_start(data = estdata,
                dtms = simple)

dtms_expectancy(probs = probs,
                dtms = simple,
                start_distr = s)
```

