---
title: "labsession1"
output:
  pdf_document: default
  html_document: default
date: "2026-07-06"
editor_options:
  chunk_output_type: console
---

## R Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# setting the wd ----------------------------------------------------------
# setting it to the course folder

setwd(file.path(here::here(),"courses","2026","DemogInR"))

# Packages
library(tidyverse)
```

## Data import

Data comes from HMD and HFD dataset for Spain in 2014.

```{r rpackages}
b <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/data/ES_B2014.csv")
d <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/data/ES_D2014.csv")
p <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/data/ES_P2014.csv")
```

## Tasks

### task 1: calculate exposures

Let's calculate the exposure based on averaging the population from the start and end of the year. Here, we are going to use this strategy.

$$E = \frac{P(1) + P(2)}{2}$$

We are going to work with tidy dataset: there is a (or some) key variables and the observation.

```{r}
e <- p |> 
  select(-open_interval) |> 
  pivot_longer(
    female:total,
    names_to = "sex",
    values_to = "pop"
  ) |> 
  # inventing column pop 1 and pop2 and taking the average based on that
  # this will work for long time series and we will do that based on lead() and lag()
  arrange(age, sex, year) |> 
  mutate(
    pop2 = lead(pop),
    .by = c(age, sex)
  ) |> 
  filter(!is.na(pop2)) |> 
  # computing exposure
  reframe(
    pop_exp = (pop + pop2) / 2,
    .by = c(age,sex)
  )
```

## task 2: calculate the fertility rates

```{r}

e |> 
  filter(sex == "total") |> 
  left_join(
    b |> select(-year, -open_interval),
    by = join_by(age)
  ) |> 
  mutate(
    total = if_else(is.na(total),0,total)
  ) |> 
  summarise(
    b = sum(total),
    e = sum(pop_exp)
  ) |> 
  mutate(
    CBR = b / e * 1000
  )

```

### exercise: age-specific fertility rates

Join to female, and calculate the rate doing by age.

```{r}
asfr <- e |> 
  filter(sex == "female") |> 
  right_join(
    b |> select(-year, -open_interval),
    by = join_by(age)
  ) |>
  mutate(
    asfr = total / pop_exp
  )

# plotting it
asfr |> 
  ggplot() +
  aes(x = age, y = asfr) +
  geom_step(linewidth = 1.05)
```

We can add up the ASFR to get the TFR:
```{r}
asfr |> 
  summarise(tfr = sum(asfr))
```

This is a synthetic cohort, and we assume that they all survived untill the end of the reproductive period.


## mortality rates

### exercise: calculate age-specific mortality rates

Let's do similarly as for fertility:

```{r}
asmr <- d |> 
  select(-year, -open_interval) |> 
  pivot_longer(
    c(female, male, total),
    names_to = "sex",
    values_to = "D"
  ) |> 
  left_join(
    e,
    by = join_by(sex, age)
  ) |> 
  # computing asmr
  mutate(mx = D / pop_exp)
```

We can calculate the CDR with this too:
```{r}
cdr <- asmr |> 
  summarise(
    CDR = sum(D) / sum(pop_exp) * 1000,
    .by = sex
  )
```

Visualizing age-specific rates

```{r}
asmr |> 
  filter(age <= 109) |> 
  ggplot() +
  aes(x = age, y = mx, color = sex) +
  geom_line(linewidth = 1.05) +
  scale_y_log10() +
  geom_hline(mapping = aes(yintercept = CDR, color = sex), data = cdr)

```

